---
title: "GARP analysis on glioblastoma (GBM)"
date: "`r Sys.Date()`"
author: Yuzhou Chang
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
---

library(rmarkdown)
```{r setup, include=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggpubr)
library(survival)
library(lubridate)
library(survminer)
library(ggplot2)
library(grid)
library(gridExtra)
library(GSVA)
```
# Analysis task
**1.	Effect of higher GARP expression (i.e. top 20% vs bottom 20%) on survival of GBM patients.**  
*Solution*: This question needs large cohort bulk-RNAseq data with clinical and treatment information. We can perform survival analysis by grouping patients by GARP expression into top 20% and bottom 20% groups.


**2.	Effect of GARP expression on the phenotype of GBM, i.e. whether there is a propensity to have a mesenchymal vs neuronal vs classical phenotype via gene signature.**  
*Solution*: This question needs large cohort bulk RNA-seq data with GBM phenotype. We can compare GARP expression or GARP-TGFb relevant pathway activity (like I did analysis for Anqi’s paper).   
The GARP-TGFb pathway relevant genes (Anqi provided):  

Activation (activation co-factors): GARP (LRRC32), ITGB6, ITGB8, ITGAV 

Platelet marker (co-factors): CD41 (ITGA2B), SELP, CD42a (GP9), CD42b (GP1BA)

Thrombin coding gene (cleave GARP): F2


**3.	Effect of GARP expression on angiogenesis, myeloid compartment, and T cell signaling/exhaustion pathways**
*Solution*: We can also use bulk-RNAseq data. First, we can compute the activity of angiogenesis, myeloid compartment, and T cell signaling/exhaustion pathways for each sample. Second, we can calculate the correlation between GARP expression and those pathways’ activity. 


# Data Overview
TCGA_cell_2013 have 152 patients and the expression data was downloaded from [cBioPortal](!https://www.cbioportal.org/datasets) and the original paper is [here](!https://pubmed.ncbi.nlm.nih.gov/24120142/)

## Preprocessing for contructing analysis object
```{r, echo=FALSE}
wd <- "e:/Dropbox/my-projects/Li_lab/Dan/"
setwd(paste0(wd,"/biocPortal/"))
#load("/bmbl_data/yuzhou/Li_lab/Dan/analysis/data_99_patient/meta_data.Rdata")

# read.pathway
angiogenesis_geneset <- read.table("../pathway_GSEA/ANGIOGENESIS.v2022.1.Hs.grp",header = T)
angiogenesis_geneset <- angiogenesis_geneset$ANGIOGENESIS
myeloid_compartment_geneset <- read.table("../pathway_GSEA/BROWN_MYELOID_CELL_DEVELOPMENT_UP.v2022.1.Hs.grp",header = T)
myeloid_compartment_geneset <- myeloid_compartment_geneset$BROWN_MYELOID_CELL_DEVELOPMENT_UP
Tcell_signal_geneset <- read.table("../pathway_GSEA/KEGG_T_CELL_RECEPTOR_SIGNALING_PATHWAY.v2022.1.Hs.grp",header = T)
Tcell_signal_geneset <- Tcell_signal_geneset$KEGG_T_CELL_RECEPTOR_SIGNALING_PATHWAY
Tcell_exhaustion_geneset <- read.table("../pathway_GSEA/GSE9650_EXHAUSTED_VS_MEMORY_CD8_TCELL_DN.v2022.1.Hs.grp",header = T)
Tcell_exhaustion_geneset <- Tcell_exhaustion_geneset$GSE9650_EXHAUSTED_VS_MEMORY_CD8_TCELL_DN

GARP_list <- c("LRRC32", "ITGB6", "ITGB8", "ITGAV",
               "ITGA2B", "SELP", "GP9", "GP1BA","F2")
pathway_list <- list(angiogenesis_geneset = angiogenesis_geneset,
                     myeloid_compartment_geneset = myeloid_compartment_geneset,
                     Tcell_signal_geneset = Tcell_signal_geneset,
                     Tcell_exhaustion_geneset = Tcell_exhaustion_geneset,
                     GARP_geneset = GARP_list)

# TCGA_cell_2013
TCGA_cell_2013_sample <- read.table("TCGA_cell_2013/data_clinical_sample.txt",header = T,check.names = F)
TCGA_cell_2013_patient <- read.delim2("TCGA_cell_2013/data_clinical_patient.txt",check.names = F,header = T,comment.char = "#")
identical(rownames(TCGA_cell_2013_sample),TCGA_cell_2013_patient$PATIENT_ID)
TCGA_cell_2013_meta <- as.data.frame(cbind(ID = TCGA_cell_2013_patient$PATIENT_ID,
                             sex = TCGA_cell_2013_patient$SEX,
                             tutmor_type = TCGA_cell_2013_sample$SAMPLE_ID,
                             OS_STATUS = TCGA_cell_2013_patient$OS_STATUS,
                             OS_time = TCGA_cell_2013_patient$DFS_MONTHS))
rownames(TCGA_cell_2013_meta) <- TCGA_cell_2013_meta$ID
TCGA_cell_2013_exp <- read.delim("TCGA_cell_2013/data_mrna_seq_v2_rsem.txt",header = T,check.names = F)
TCGA_cell_2013_exp <- TCGA_cell_2013_exp[-which(duplicated(TCGA_cell_2013_exp$Hugo_Symbol)),]
rownames(TCGA_cell_2013_exp) <- TCGA_cell_2013_exp$Hugo_Symbol
TCGA_cell_2013_exp$Hugo_Symbol <- NULL
TCGA_cell_2013_exp$Entrez_Gene_Id <- NULL
colnames(TCGA_cell_2013_exp) <- gsub("-01$","",colnames(TCGA_cell_2013_exp))
TCGA_cell_2013_meta <- TCGA_cell_2013_meta[colnames(TCGA_cell_2013_exp),]
GARP_list %in% rownames(TCGA_cell_2013_exp)
# run pathway
TCGA_cell_2013_pathway <- gsva(expr = as.matrix(TCGA_cell_2013_exp), gset.idx.list = pathway_list, verbose=FALSE)
TCGA_cell_2013_meta <- TCGA_cell_2013_meta[colnames(TCGA_cell_2013_pathway),]
identical(rownames(TCGA_cell_2013_meta),colnames(TCGA_cell_2013_pathway))
identical(rownames(TCGA_cell_2013_meta),colnames(TCGA_cell_2013_exp))
TCGA_cell_2013_meta$angiogenesis_geneset <- TCGA_cell_2013_pathway["angiogenesis_geneset",]
TCGA_cell_2013_meta$myeloid_compartment_geneset <- TCGA_cell_2013_pathway["myeloid_compartment_geneset",]
TCGA_cell_2013_meta$Tcell_signal_geneset <- TCGA_cell_2013_pathway["Tcell_signal_geneset",]
TCGA_cell_2013_meta$Tcell_exhaustion_geneset <- TCGA_cell_2013_pathway["Tcell_exhaustion_geneset",]
TCGA_cell_2013_meta$GARP_geneset <- TCGA_cell_2013_pathway["GARP_geneset",]
TCGA_cell_2013_meta$GARP_expression <- as.numeric(TCGA_cell_2013_exp["LRRC32",])

```


# Task 1: Effect of higher GARP expression (i.e. top 20% vs bottom 20%) on survival of GBM patients.
## Task 1 - TCGA_cell_2013 survival analysis for all tumor subtypes
```{r, echo = F}
# create censor
TCGA_cell_2013_meta$censor <- ifelse(TCGA_cell_2013_meta$OS_STATUS == "1:DECEASED",1,0)
# TCGA_cell_2013_meta[,c(2,3)] %>% tbl_summary()
TCGA_cell_2013_meta <- TCGA_cell_2013_meta[TCGA_cell_2013_meta$tutmor_type != "G-CIMP" & !is.na(TCGA_cell_2013_meta$tutmor_type),]
#
TCGA_cell_2013_meta$GARP_binary <- "no_group"
TCGA_cell_2013_meta$GARP_binary <- ifelse(TCGA_cell_2013_meta$GARP_expression > quantile(TCGA_cell_2013_meta$GARP_expression,0.8,na.rm = T ),"high_expression",TCGA_cell_2013_meta$GARP_binary)
TCGA_cell_2013_meta$GARP_binary <- ifelse(TCGA_cell_2013_meta$GARP_expression < quantile(TCGA_cell_2013_meta$GARP_expression,0.2,na.rm = T),"low_expression",TCGA_cell_2013_meta$GARP_binary)
# table(TCGA_cell_2013_meta$GARP_binary )
sub.meta.GARP <- TCGA_cell_2013_meta[TCGA_cell_2013_meta$GARP_binary!="no_group",]
sub.meta.GARP$OS_time <- as.numeric(sub.meta.GARP$OS_time)
km_trt_fit <- survfit(Surv(OS_time, censor) ~ GARP_binary, data=sub.meta.GARP)
ggsurvplot(km_trt_fit,pval = T)

```

## Task 1 - TCGA_cell_2013 survival analysis for four tumor subtypes 

### Survival analysis for classical
```{r, echo =F}
TCGA_cell_2013_meta_Classical <- TCGA_cell_2013_meta[TCGA_cell_2013_meta$tutmor_type == "Classical",]
sub.meta_Classical.GARP <- TCGA_cell_2013_meta_Classical[TCGA_cell_2013_meta_Classical$GARP_binary!="no_group",]
sub.meta_Classical.GARP$OS_time <- as.numeric(sub.meta_Classical.GARP$OS_time)
km_trt_fit <- survfit(Surv(OS_time, censor) ~ GARP_binary, data=sub.meta_Classical.GARP)
ggsurvplot(km_trt_fit,pval = T) + labs(title = "classical") 
```

### Survival analysis for mesenchymal  
```{r, echo = F}
TCGA_cell_2013_meta_Mesenchymal <- TCGA_cell_2013_meta[TCGA_cell_2013_meta$tutmor_type == "Mesenchymal",]
sub.meta_Mesenchymal.GARP <- TCGA_cell_2013_meta_Mesenchymal[TCGA_cell_2013_meta_Mesenchymal$GARP_binary!="no_group",]
sub.meta_Mesenchymal.GARP$OS_time <- as.numeric(sub.meta_Mesenchymal.GARP$OS_time)
km_trt_fit <- survfit(Surv(OS_time, censor) ~ GARP_binary, data=sub.meta_Mesenchymal.GARP)
ggsurvplot(km_trt_fit,pval = T, ggtheme = theme_classic())+ labs(title = "Mesenchymal")
```

### Survival analysis for  neural  
```{r, echo = F}
TCGA_cell_2013_meta_Neural <- TCGA_cell_2013_meta[TCGA_cell_2013_meta$tutmor_type == "Neural",]
sub.meta_Neural.GARP <- TCGA_cell_2013_meta_Neural[TCGA_cell_2013_meta_Neural$GARP_binary!="no_group",]
sub.meta_Neural.GARP$OS_time <- as.numeric(sub.meta_Neural.GARP$OS_time)
km_trt_fit <- survfit(Surv(OS_time, censor) ~ GARP_binary, data=sub.meta_Neural.GARP)
ggsurvplot(km_trt_fit,pval = T)+ labs(title = "Neural") 
```

### Survival analysis proneural  
```{r, echo = F}
TCGA_cell_2013_meta_Proneural <- TCGA_cell_2013_meta[TCGA_cell_2013_meta$tutmor_type == "Proneural",]
sub.meta_Proneural.GARP <- TCGA_cell_2013_meta_Proneural[TCGA_cell_2013_meta_Proneural$GARP_binary!="no_group",]
sub.meta_Proneural.GARP$OS_time <- as.numeric(sub.meta_Proneural.GARP$OS_time)
km_trt_fit <- survfit(Surv(OS_time, censor) ~ GARP_binary, data=sub.meta_Proneural.GARP)
ggsurvplot(km_trt_fit,pval = T)+ labs(title = "Proneural") 

```


# Task 2:	Effect of GARP expression on the phenotype of GBM
In this analysis, we will compare GARP expression and GARP pathway activity among different GBM subtypes. The pathway activity is calculated by the GSVA algorithm.

## Task 2-TCGA_cell_2013 

We calculate p-values for any two groups regarding GARP expression and GARP pathway activity. 

* Classical vs Mesenchymal 
* Classical vs Proneural
* Classical vs Neural
* Mesenchymal vs Proneural
* Mesenchymal vs Neural
* Proneural vs Neural

###  GARP expression

```{r, echo = F,warning=F}
TCGA_cell_2013_meta_GARP <-TCGA_cell_2013_meta[,c("tutmor_type","GARP_expression")]
TCGA_cell_2013_meta_GARP <- TCGA_cell_2013_meta_GARP[!is.na(TCGA_cell_2013_meta_GARP$GARP_expression),]
my.comparison <- list(c("Classical","Mesenchymal"),
                      c("Classical", "Proneural"),
                      c("Classical","Neural"),
                      c("Mesenchymal","Proneural"),
                      c("Mesenchymal","Neural"),
                      c("Neural","Proneural"))



ggboxplot(TCGA_cell_2013_meta_GARP, x = "tutmor_type", y = "GARP_expression",
          color = "tutmor_type", palette = "jco")+ 
  stat_compare_means(comparisons = my.comparison)

```

### GARP pathway activity

```{r, echo = F,warning=F}
TCGA_cell_2013_meta_GARP <-TCGA_cell_2013_meta[,c("tutmor_type","GARP_geneset")]
TCGA_cell_2013_meta_GARP <- TCGA_cell_2013_meta_GARP[!is.na(TCGA_cell_2013_meta_GARP$GARP_geneset),]
my.comparison <- list(c("Classical","Mesenchymal"),
                      c("Classical", "Proneural"),
                      c("Classical","Neural"),
                      c("Mesenchymal","Proneural"),
                      c("Mesenchymal","Neural"),
                      c("Neural","Proneural"))



ggboxplot(TCGA_cell_2013_meta_GARP, x = "tutmor_type", y = "GARP_geneset",
          color = "tutmor_type", palette = "jco")+ 
  stat_compare_means(comparisons = my.comparison)

```

# Task 3:	Effect of GARP expression and GARP pathway on angiogenesis, myeloid compartment, and T cell signaling/exhaustion pathways

## Task3-TCGA_cell_2013

### GARP Pearson correlation results based on all patients

The X-axis is the GARP expression value. 

The Y-axis is the enrichment score (a higher score means higher activity) regarding four pathways.

```{r,echo=F,message=FALSE,warning=F}
TCGA_cell_2013_meta$GARP_expression_log <- log1p(TCGA_cell_2013_meta$GARP_expression)
p.angio <- ggscatter(TCGA_cell_2013_meta, x = "GARP_expression_log", y = "angiogenesis_geneset",
          add = "reg.line",  size = 1,                                # Add regression line
          conf.int = TRUE,                                  # Add confidence interval
          add.params = list(color = "blue",
                            fill = "lightgray")
          )+
  stat_cor(method = "pearson") + labs(title = "GARP vs angiogenesis",x = "log(GARP)", y ="angiogenesis enrichment score")+
  theme(axis.text = element_text(size = 10),text = element_text(size = 10)) # Add correlation coefficient

p.myeloid <- ggscatter(TCGA_cell_2013_meta, x = "GARP_expression_log", y = "myeloid_compartment_geneset",
          add = "reg.line",  size = 1,                                # Add regression line
          conf.int = TRUE,                                  # Add confidence interval
          add.params = list(color = "blue",
                            fill = "lightgray")
          )+
  stat_cor(method = "pearson") + labs(title = "GARP vs myeloid_compartment",x = "log(GARP)", y ="myeloid compartment enrichment score") +
  theme(axis.text = element_text(size = 10),text = element_text(size = 10))# Add correlation coefficient

p.Tcell_signal <- ggscatter(TCGA_cell_2013_meta, x = "GARP_expression_log", y = "Tcell_signal_geneset",
          add = "reg.line", size = 1,                                 # Add regression line
          conf.int = TRUE,                                  # Add confidence interval
          add.params = list(color = "blue",
                            fill = "lightgray")
          )+
  stat_cor(method = "pearson") + labs(title = "GARP vs Tcell_signal",x = "log(GARP)", y ="Tcell_signal enrichment score")+
  theme(axis.text = element_text(size = 10),text = element_text(size = 10)) # Add correlation coefficient

p.Tcell_exhaustion <- ggscatter(TCGA_cell_2013_meta, x = "GARP_expression_log", y = "Tcell_exhaustion_geneset",
          add = "reg.line", size = 1,                                 # Add regression line
          conf.int = TRUE,                                  # Add confidence interval
          add.params = list(color = "blue",
                            fill = "lightgray")
          )+
  stat_cor(method = "pearson") + labs(title = "GARP vs Tcell_exhaustion",x = "log(GARP)", y ="Tcell_exhaustion nrichment score")+
  theme(axis.text = element_text(size = 10),text = element_text(size = 10)) # Add correlation coefficient

grid.arrange(p.angio, p.myeloid, p.Tcell_signal,p.Tcell_exhaustion )

```

### GARP pathway Pearson correlation results based on all patients

The X-axis is the GARP pathway enrichment score (a higher score means higher activity). 

The Y-axis is the enrichment score regarding four pathways.

```{r,echo=F,message=FALSE,warning=F}

p.angio <- ggscatter(TCGA_cell_2013_meta, x = "GARP_geneset", y = "angiogenesis_geneset",
          add = "reg.line",  size = 1,                                # Add regression line
          conf.int = TRUE,                                  # Add confidence interval
          add.params = list(color = "blue",
                            fill = "lightgray")
          )+
  stat_cor(method = "pearson") + labs(title = "GARP pathway vs angiogenesis",x = "GARP pathway enrichment score", y ="angiogenesis enrichment score") +
  theme(axis.text = element_text(size = 10),text = element_text(size = 10))# Add correlation coefficient

p.myeloid <- ggscatter(TCGA_cell_2013_meta, x = "GARP_geneset", y = "myeloid_compartment_geneset",
          add = "reg.line",  size = 1,                                # Add regression line
          conf.int = TRUE,                                  # Add confidence interval
          add.params = list(color = "blue",
                            fill = "lightgray")
          )+
  stat_cor(method = "pearson") + labs(title = "GARP pathway vs myeloid_compartment",x = "GARP pathway enrichment score", y ="myeloid compartment enrichment score") +
  theme(axis.text = element_text(size = 10),text = element_text(size = 10))# Add correlation coefficient

p.Tcell_signal <- ggscatter(TCGA_cell_2013_meta, x = "GARP_geneset", y = "Tcell_signal_geneset",
          add = "reg.line",  size = 1,                                # Add regression line
          conf.int = TRUE,                                  # Add confidence interval
          add.params = list(color = "blue",
                            fill = "lightgray")
          )+
  stat_cor(method = "pearson") + labs(title = "GARP pathway vs Tcell_signal",x = "GARP pathway enrichment score", y ="Tcell_signal enrichment score")+
  theme(axis.text = element_text(size = 10),text = element_text(size = 10)) # Add correlation coefficient

p.Tcell_exhaustion <- ggscatter(TCGA_cell_2013_meta, x = "GARP_geneset", y = "Tcell_exhaustion_geneset",
          add = "reg.line",  size = 1,                                # Add regression line
          conf.int = TRUE,                                  # Add confidence interval
          add.params = list(color = "blue",
                            fill = "lightgray")
          )+
  stat_cor(method = "pearson") + labs(title = "GARP pathway vs Tcell_exhaustion",x = "GARP pathway enrichment score", y ="Tcell_exhaustion nrichment score")+
  theme(axis.text = element_text(size = 10),text = element_text(size = 10)) # Add correlation coefficient

grid.arrange(p.angio, p.myeloid, p.Tcell_signal,p.Tcell_exhaustion )
```






